Supporting Information: Unsupervised embedding of trajectories captures the latent structure of mobility

### 2023-08-24 paper/current/nature-comm-physics/supp.tex
many small changes

In the case that an individual lists multiple affiliations in a single year, affiliations listed on publications published in that year are shuffled between each iteration of the \textit{word2vec} training process (each epoch).
For example, an author who published with affiliation $A$ in $t_{0}$, and affiliations $B$ and $C$ in $t_{1}$ could appear as one of $(A, B, C)$ or $(A, C, B)$ in each training iteration.
This effectively removes the effect of order within a year, as the order cannot be mFigeaningfully established based on co-affiliations in a single paper, or on different affiliations listed on separate papers, for which the date of publication may not be representative of the actual completion of the project.

@@

We start by generating synthetic mobility trajectories structured according the gravity model. For this purpose, we first generate an artificial ``geographic'' space. 1,000 points are generated in a 2-dimensional space, with their coordinates distributed evenly between -1 and 1. These points act as the ``locations'' in our simulation. Each point is also assigned a ``mass'' randomly drawn from a power-law distribution $p(x) \propto x^{-k}$ (we use $k=2$) that is meant to represent the ``population'' or ``size'' of each location.

With this simulated geography in place, we calculate the pairwise distance matrix $r_{ij}^{synthetic}$, which is then used to calculate a transition matrix between points, defined as $T_{ij} = M_i M_j f(r_{ij})$, with transition probabilities following the gravity law. The purpose of this transition matrix is to define the likelihood of movement between points (or \textit{locations}) in our simulated geography based on their mass and pairwise distances. We conduct this experiment with two different transition matrices, one based on cosine distance where $f(r_{ij})=e^{-\beta r_{ij}}$ and another with euclidean distance where $f(r_{ij})=r_{ij}^{-\alpha}$. For simplicity, we use $\beta=2$ and $\alpha=2$.

@@

From the embedding trained with synthetic trajectories, we calculate the pairwise embedding distances between points, $r_{ij}^{embedding}$.
To evaluate the ability \textit{word2vec} to recover the known gravity-like structure of the synthetic data, we compare the embedding distance to the distances between the 1,000 points in the synthetic space, $r_{ij}^{synthetic}$ using the Pearson correlation coefficient. 
This result is repeated 100 times, and the correlation coefficients are aggregated across iterations.

We find that \textit{word2vec} effectively recovers the gravity-like relationships in the synthetic data. 
Based on cosine distance, the Pearson correlation between the embedding and synthetic distances is 0.943, with a standard error of 0.01. 
Based on euclidean distance, the correlation was 0.801, with a standard error of 0.01.
To better illustrate this point, Fig. \ref{fig:supp:synthetic_exp} shows the correlation for one representative iteration of this experiment.

In order to evaluate the robustness of the outcome across hyperparameter settings, we repeat our synthetic benchmark (using cosine distance) varying values for the mass exponent, $k$ and distance exponent $\beta$. 
In Fig. \ref{fig:supp:synthetic_robust}) presents the average Pearson correlation between the synthetic and embedding distance over ten iterations with for different values of $k$ and $\beta$. 
This shows our findings to be robust. 
Neural embedding consistently recovers the distance from the synthetic trajectories across a broad range of hyperparameter values. 
A low correlation occurs only when the mass distribution is too heterogeneous ($k < 1.9$).
In a heterogeneous mass distribution, most of the trajectories will concentrate on just a few high-mass locations, regardless of distance, which makes learning a distance challenging.  Meanwhile, the exponent of the city population usually ranges between 2 and 3~\autocite{clauset2009power}.

>>>>

In the case that an individual lists multiple affiliations in a single year, affiliations listed on publications published in that year are shuffled between each iteration of the \textit{word2vec} training process (each epoch).
For example, an author who published with affiliation $A$ in $t_{0}$, and affiliations $B$ and $C$ in $t_{1}$ could appear as one of $(A, B, C)$ or $(A, C, B)$ in each training iteration.
This effectively removes the effect of order within a year, as the order cannot be meaningfully established based on co-affiliations in a single paper, or on different affiliations listed on separate papers, for which the date of publication may not be representative of the actual completion of the project.

@@

We start by generating synthetic mobility trajectories structured according to the gravity model. For this purpose, we first generate an artificial ``geographic'' space. 1,000 points are generated in a 2-dimensional space, with their coordinates distributed evenly between -1 and 1. These points act as the ``locations'' in our simulation. Each point is also assigned a ``mass'' randomly drawn from a power-law distribution $p(x) \propto x^{-k}$ (we use $k=2$) that is meant to represent the ``population'' or ``size'' of each location.

With this simulated geography in place, we calculate the pairwise distance matrix $r_{ij}^{synthetic}$, which is then used to calculate a transition matrix between points, defined as $T_{ij} = M_i M_j f(r_{ij})$, with transition probabilities following the gravity law. The purpose of this transition matrix is to define the likelihood of movement between points (or \textit{locations}) in our simulated geography based on their mass and pairwise distances. We conduct this experiment with two different transition matrices, one based on cosine distance where $f(r_{ij})=e^{-\beta r_{ij}}$ and another with Euclidean distance where $f(r_{ij})=r_{ij}^{-\alpha}$. For simplicity, we use $\beta=2$ and $\alpha=2$.

@@

From the embedding trained with synthetic trajectories, we calculate the pairwise embedding distances between points, $r_{ij}^{embedding}$.
To evaluate the ability of \textit{word2vec} to recover the known gravity-like structure of the synthetic data, we compare the embedding distance to the distances between the 1,000 points in the synthetic space, $r_{ij}^{synthetic}$ using the Pearson correlation coefficient. 
This result is repeated 100 times, and the correlation coefficients are aggregated across iterations.

We find that \textit{word2vec} effectively recovers the gravity-like relationships in the synthetic data. 
Based on cosine distance, the Pearson correlation between the embedding and synthetic distances is 0.943, with a standard error of 0.01. 
Based on Euclidean distance, the correlation was 0.801, with a standard error of 0.01.
To better illustrate this point, Fig. \ref{fig:supp:synthetic_exp} shows the correlation for one representative iteration of this experiment.

In order to evaluate the robustness of the outcome across hyperparameter settings, we repeat our synthetic benchmark (using cosine distance) varying values for the mass exponent, $k$ and distance exponent $\beta$. 
Fig. \ref{fig:supp:synthetic_robust} presents the average Pearson correlation between the synthetic and embedding distance over ten iterations with for different values of $k$ and $\beta$. 
This shows our findings to be robust. 
Neural embedding consistently recovers the distance from the synthetic trajectories across a broad range of hyperparameter values. 
A low correlation occurs only when the mass distribution is too heterogeneous ($k < 1.9$).
In a heterogeneous mass distribution, most of the trajectories will concentrate on just a few high-mass locations, regardless of distance, which makes learning a distance challenging.  Meanwhile, the exponent of the city population usually ranges between 2 and 3~\autocite{clauset2009power}.

### 2023-08-24 paper/current/pnas/main.tex
many small changes

% Math
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{newtxmath}
\DeclareMathAlphabet{\mathpzc}{T1}{pzc}{m}{it}
\def\given{\mid}
\usepackage{bm}
\def\tnull{{\text{null}}}
\def\vec#1{{\bm #1}}
\def\mat#1{\mathbf{#1}}

@@

% for quick author comments
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{light-gray}{gray}{0.8}
\def\del#1{ {\color{light-gray}{#1}} }
\def\yy#1{ {\color{red}\textbf{yy: #1}} }
\def\dk#1{ {\color{red}\textbf{dk: #1}} }
\def\js#1{ {\color{red}\textbf{js: #1}} }
\def\sk#1{ {\color{red}\textbf{sk: #1}} }

@@

@@

%
% Figure - Boomerang
%
\begin{figure}[h!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/Math/boomerang.pdf}
	\caption{
		\textbf{Size of organization embedding vectors captures prestige and size of organizations.}
		\textbf{a.} Size (L2 norm) of organization embedding vectors compared to the number of researchers for U.S. universities.
		Color indicates the rank of the university from the Times ranking, with 1 being the highest ranked university.
		Uncolored points are universities not listed on the Times ranking.
		A concave-shape emerges, wherein larger universities tend to be more distant from the origin (large L2 norm); however, the more prestigious universities tend to have smaller L2 norms.
		\textbf{b.} We find a similar concave-curve pattern across many countries such as the United States, China, Australia, Brazil, and others (\textbf{inset}, and Fig.~\ref{fig:concave30}).
		Some countries exhibit variants of this pattern, such as Egypt, which is missing the right side of the curve.
		The loess regression lines are shown for each selected country, and for the aggregate of remaining countries, with ribbons mapping to the 99\% confidence intervals based on a normal distribution.
		Loess lines are also shown for organizations in Australia, Brazil, and Egypt (\textbf{inset}).
	}
	\label{fig:length}
\end{figure}

>>>>

% Math
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{newtxtext,newtxmath}
\DeclareMathAlphabet{\mathpzc}{T1}{pzc}{m}{it}
\def\given{\mid}
\usepackage{bm}
\def\tnull{{\text{null}}}
\def\vec#1{{\bm #1}}
\def\mat#1{\mathbf{#1}}

@@

@@

Just as the embedding space makes it possible to \textit{zoom in} on subsets of organizations, it is also possible to \textit{zoom out} by aggregating organizational vectors.
In doing so, we identify distinct clusters of countries, organized by their language, culture, and geographic proximity.
Moreover, we find that

@@

### 2023-08-24 paper/current/pnas/supp.tex
many small changes

%
% S14 TEXT
\paragraph*{S15 Text}
\label{si:text:boomerang_speculation}
{\bf Speculation on variations of the convex-curve pattern.}

@@

>>>>

%
% S14 TEXT
\paragraph*{S15 Text}
\label{si:text:boomerang_speculation}
{\bf A convex-curve pattern in organization prestige.}
We also find that the size (L2 norm) of the organization embedding vectors provides insights into the characteristics of organizations (Fig.~\ref{fig:length}).
Up to a point (around 1,000 researchers), the size of U.S. organization's vectors tends to increase proportionally to the number of researchers (both mobile and non-mobile) with published work;
these organizations are primarily teaching-focused institutions, agencies, and hospitals that either are not ranked or have a low ranking.
However, at around 1,000 researchers, the size of the vector \emph{decreases} as the number of researchers increases.
These organizations are primarily research-intensive and prestigious universities with higher rank, research outputs, R\&D funding, and doctoral students (Fig.~\ref{fig:supp:length_vs_metaInfo}).
A similar pattern has been observed in applications of neural embedding to natural language, in which the size of word vectors were found to represent the word's \emph{specificity}, i.e., the word associated with the vector frequently co-appears with particular context words\autocite{schakel2015measuring}.
If the word in question is universal, appearing frequently in many different contexts, it would not have a large norm due to a lack of strong association with a particular context.
Likewise, an organization with a small norm, such as Harvard, appears in many contexts alongside many different organizations in affiliation trajectories---it is well-connected.
The concavity of the curve emerges in part from the relationship between the size of the vector and the expected connectedness of the organization, given its size ($R^{2} = 0.17$).
Large, prestigious, and well-funded research universities such as Princeton and Harvard have smaller vector norms because they appear in many different contexts compared to more teaching-focused organizations such as NY Medical College, and the University of Michigan at Flint.
Some universities, such as the University of Alaska at Fairbanks, have considerably small vectors, which may be a result of their remote locations and unique circumstances.

We report that this curve is almost universal across many countries.
For instance, China's curve closely mirrors that of the United States (Fig.~\ref{fig:length}b).
Smaller but scientifically advanced countries such as Australia and other populous countries such as Brazil also exhibit curves similar to the United States (Fig.~\ref{fig:length}b, inset).
Other nations exhibit different curves which lack the portions with decreasing norm, probably indicating the lack of internationally-prestigious institutions.
Similar patterns can be found across many of the 30 countries with the most total researchers (Fig.~\ref{fig:concave30}).  Although we speculate on potential explanations for this pattern below, we don't intend to make general claims about scientific mobility, or mobility more generally, but rather to illustrate how \textit{word2vec} can facilitate new kinds of investigations into mobility.
Further work is necessary to determine what causes this curve pattern in so many countries and whether the same pattern can be found in other domains of human mobility.

@@

%
% Figure - Boomerang
%
\begin{figure}[h!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/Math/boomerang.pdf}
	\caption{
		\textbf{Size of organization embedding vectors captures prestige and size of organizations.}
		\textbf{a.} Size (L2 norm) of organization embedding vectors compared to the number of researchers for U.S. universities.
		Color indicates the rank of the university from the Times ranking, with 1 being the highest ranked university.
		Uncolored points are universities not listed on the Times ranking.
		A concave-shape emerges, wherein larger universities tend to be more distant from the origin (large L2 norm); however, the more prestigious universities tend to have smaller L2 norms.
		\textbf{b.} We find a similar concave-curve pattern across many countries such as the United States, China, Australia, Brazil, and others (\textbf{inset}, and Fig.~\ref{fig:concave30}).
		Some countries exhibit variants of this pattern, such as Egypt, which is missing the right side of the curve.
		The loess regression lines are shown for each selected country, and for the aggregate of remaining countries, with ribbons mapping to the 99\% confidence intervals based on a normal distribution.
		Loess lines are also shown for organizations in Australia, Brazil, and Egypt (\textbf{inset}).
	}
	\label{fig:length}
\end{figure}
